### 2019-07-07 1-Preamble/5-Abstract/resume_fr.tex
De meilleures transitions, j'ai retravaillé chaque abstract

Suite à deux travaux précédents récents \cite{LiuLeeShroff17,CaoZhenKvetonXie18}, nous combinons une politique d'index efficace (\klUCB) et un test efficace de détection des points de changement, dans l'hypothèse de récompenses limitées.
Nous nous appuyons sur des résultats très récents du test du rapport de vraisemblance généralisé (TRGL) pour les variables gaussiennes et sous-gaussiennes \cite{Maillard2018GLR}, mais nous nous concentrons plutôt sur les récompenses limitées et les distributions de Bernoulli.
Les récompenses limitées sont en effet généralement plus appropriées pour les applications de radio cognitive, comme indiqué ci-dessus.
En utilisant le fait que les variables bornées en $[0,1]$ ne sont pas seulement sous-Gaussiennes à $1/4$ mais aussi sous-Bernoulli, nous prouvons les premières garanties à temps fini pour le GLRT pour les variables bornées.
Nous montrons d'abord des limites temporelles finies à la fois sur la probabilité de fausse alarme de notre test, et sur son délai de détection, sous de légères hypothèses sur la longueur des séquences stationnaires.
Notre algorithme, noté \GLRklUCB, est alors présenté en deux variantes, que les points de changement soient locaux (\ie, une seule moyenne de bras change à chaque point de changement) ou globaux (\ie, peut-être tous les moyens de bras changent à la fois).
%
Nous prouvons qu'en combinant l'algorithme \klUCB, asymptotiquement optimale pour les problèmes stationnaires, et notre nouvelle analyse du GLRT pour les récompenses limitées, nous obtenons des garanties de pointe sur le regret de notre algorithme proposé.
Nous considérons la même hypothèse que nos concurrents, en supposant que la longueur des intervalles stationnaires est " assez longue " pour que les points de changement soient détectables.
La limite supérieure du meilleur regret est obtenue lorsque l'algorithme connaît à l'avance l'horizon et le nombre de points de changement, mais notre algorithme n'a pas besoin d'avoir d'autres connaissances sur la difficulté du problème pour être accordé de manière optimale.
%
La performance de \GLRklUCB{} est également illustrée par des expériences numériques sur des données synthétiques, où il est démontré qu'elle surpasse toutes les algorithms passivement adaptatifs ainsi que les précédentes algorithms activement adaptatifs.
%
Ce dernier chapitre est basé sur notre dernier travail, publié à la conférence GRETSI à Lille, en août $2019$ \cite{Besson2019Gretsi}.
Notre travail a aussi mené à l'article de la version longue \cite{Besson2019GLRT}, qui sera partiellement réécrit et complété avec des résultats plus récents, afin de le soumettre bientôt à une revue (probablement le Journal of Machine Learning Research) avant décembre $2019$.

@@

    \item
    Nous donnons également une revue détaillée de la littérature sur les différentes extensions du modèle BMB multi-joueurs, qui a été particulièrement active ces deux dernières années.
    % qui, selon nous, n'a jamais été écrit auparavant.
    % (ALT 2018, et 4-8 travaux inspirés de notre article depuis lors). Etat de l'art avec notre algorithme MCTopM + klUCB pour les bandits multi-joueurs ``avec détection'', état de l'art empirique avec notre approche simple (mais erronée) ``égoïste'' en cas de ``non détection''.

@@

>>>>

Suite à deux travaux précédents récents \cite{LiuLeeShroff17,CaoZhenKvetonXie18}, nous combinons une politique d'index efficace (\klUCB) et un test efficace de détection des points de changement, dans l'hypothèse de récompenses limitées.
Nous nous appuyons sur des résultats très récents du test du rapport de vraisemblance généralisé (TRGL) pour les variables gaussiennes et sous-gaussiennes \cite{Maillard2018GLR}, mais nous nous concentrons plutôt sur les récompenses limitées et les distributions de Bernoulli.
Les récompenses limitées sont en effet généralement plus appropriées pour les applications de radio cognitive, comme indiqué ci-dessus.
En utilisant le fait que les variables bornées en $[0,1]$ ne sont pas seulement sous-Gaussiennes à $1/4$ mais aussi sous-Bernoulli, nous prouvons les premières garanties à temps fini pour le GLRT pour les variables bornées.
Nous montrons d'abord des limites temporelles finies à la fois sur la probabilité de fausse alarme de notre test, et sur son délai de détection, sous de légères hypothèses sur la longueur des séquences stationnaires.
Notre algorithme, noté \GLRklUCB, est alors présenté en deux variantes, que les points de changement soient locaux (\ie, une seule moyenne de bras change à chaque point de changement) ou globaux (\ie, peut-être tous les moyens de bras changent à la fois).
%
Nous prouvons qu'en combinant l'algorithme \klUCB, asymptotiquement optimale pour les problèmes stationnaires, et notre nouvelle analyse du GLRT pour les récompenses limitées, nous obtenons des garanties de pointe sur le regret de notre algorithme proposé.
Nous considérons la même hypothèse que nos concurrents, en supposant que la longueur des intervalles stationnaires est " assez longue " pour que les points de changement soient détectables.
La limite supérieure du meilleur regret est obtenue lorsque l'algorithme connaît à l'avance l'horizon et le nombre de points de changement, mais notre algorithme n'a pas besoin d'avoir d'autres connaissances sur la difficulté du problème pour être accordé de manière optimale.
%
La performance de \GLRklUCB{} est également illustrée par des expériences numériques sur des données synthétiques, où il est démontré qu'elle surpasse toutes les algorithms passivement adaptatifs ainsi que les précédentes algorithms activement adaptatifs.
%
Ce dernier chapitre est basé sur notre dernier travail, publié à la conférence GRETSI à Lille, en août $2019$ \cite{Besson2019Gretsi}.
Notre travail a aussi mené à une version longue \cite{Besson2019GLRT}, qui sera complétée avec des résultats plus récents, afin de le soumettre bientôt à une revue (probablement le Journal of Machine Learning Research) avant décembre $2019$.

@@

    \item
    Nous donnons également une revue détaillée de la littérature sur les différentes extensions du modèle BMB multi-joueurs, particulièrement active ces deux dernières années.
    % qui, selon nous, n'a jamais été écrit auparavant.
    % (ALT 2018, et 4-8 travaux inspirés de notre article depuis lors). Etat de l'art avec notre algorithme MCTopM + klUCB pour les bandits multi-joueurs ``avec détection'', état de l'art empirique avec notre approche simple (mais erronée) ``égoïste'' en cas de ``non détection''.

@@

\hr{}

### 2019-07-07 2-Chapters/1-Chapter/chapter1.tex
De meilleures transitions, j'ai retravaillé chaque abstract

We illustrate below the idea of a \emph{learning cycle}, alternating between action and feedback,
in Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then by observing a reward $r(t)$, which is a certain measure of success of this action, produced by the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its rewards, by trials and errors (\ie, actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk or to drive, learning to play a board or computer game, or discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

\paragraph{Opportunistic Spectrum Access.}
%
The two communities of OSA and MAB have started to interact, and the resulting works have received a great interest from both communities, since the late $2000$s and early $2010$s, with pioneers works like \cite{Liu08,Zhao10,Jouini09,Jouini10}.
The focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypotheses are maid:
the SU is equipped with a sensing capability,
and considers a fixed and finite set of $K$ orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of $K=3$ Wi-Fi channels at different frequency, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in time, by sub-dividing the time in discrete time steps (or rounds).
%
Thus if the SU spends a short time in the beginning of each time slot to perform sensing, it can scan for the presence or absence of any PU before transmitting, during the rest of the time slot without colliding with the PU.
If the SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be costly, in terms of energy consumption, especially for wide-band sensing, as detailed in the surveys \cite{yucek2009survey,subhedar2011spectrum}, thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time.
This assumption, along with the hypothesis that PU cannot be disturbed (without which no regulation will never be accepted for OSA), enforces the SU to transmit in the channel that it sensed, in each time step, if and only if it was sensed free from any PU.
% non negligible cost of radio reconfiguration

Focussing on one SU in an OSA network, it has to sequentially decide a channel to sense (in the set of channels, $[K]=\{1,\dots,K\}$), then it performs sensing, and finally it transmits in this channel if it was sensed free.
The goal of the SU is to minimize its energy consumption (we remind that we focus on ``green'' cognitive radio) and to maximize its uplink data rate, or equivalently, to maximize its number of successful transmission.
%
If the different channels are not uniformly used by the PU, and if we assume a stationary hypothesis on the PU traffic, then the goal of the SU boils down to exploring the different channels and exploiting the best ones.
This frames the Opportunistic Spectrum Access problem as an exploration/exploitation problem,
with a finite set of actions (the channels, also called \emph{arms}),
in a sequential action-then-feedback cycle (time steps are $t\in\N^*$),
under partial information feedback (\ie, the player receives feedback about only one channel over $K$).
These three hypotheses are the ones that restrict from the general ML framework to the specific Multi-Armed Bandit case (see Figure~\ref{fig:1:ReinforcementLearningCycle}).

@@

We start by formulating the problem studied in this thesis, and then we develop the organization of the contributions.
%
% \textbf{Problematic.}
%
If we sum-up the studied problems and summarize them in one question, it could be the following:
\emph{``Can we adapt the decision making tools already successfully applied to Cognitive Radio for Opportunistic Spectrum Access to the specific needs of CR for the (future) Internet of Things networks?''}
%
We answer partially to this problematic by the following research steps.

@@

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core concept is to let every dynamic device of such network run a learning algorithm in a fully decentralized way, in order to optimize the complete system.
It is important to note that each devices communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
It means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic to the other devices following the same objective.
It is well known in the game theory literature that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning lead to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

@@

\item
    \emph{What Doubling-Trick Can and Can't Do for Multi-Armed Bandits},
    by \textbf{Lilian Besson} \& Emilie Kaufmann,
    September $2018$.\\
    % See Appendix~\ref{app:2:DoublingTricks}.
    Preprint at \href{https://HAL.Inria.fr/hal-01736357}{\texttt{HAL.Inria.fr/hal-01736357}}.
    \cite{Besson2018DoublingTricks}

>>>>

We illustrate below the idea of a \emph{learning cycle}, alternating between action and feedback,
in Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then by observing a reward $r(t)$, which is a certain measure of success of this action, produced by the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its rewards, by trials and errors (\ie, actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk or drive, to play a board or computer game, or discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

\paragraph{Opportunistic Spectrum Access.}
%
The two communities of OSA and MAB have started to interact, and the resulting works have received a great interest from both communities, since the late $2000$s and early $2010$s, with pioneers works like \cite{Liu08,Zhao10,Jouini09,Jouini10}.
The focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypotheses are maid:
the SU is equipped with a sensing capability,
and considers a fixed and finite set of $K$ orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of $K=3$ Wi-Fi channels at different frequency, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in time, by sub-dividing the time in discrete time steps (or rounds).
%
Thus if the SU spends a short time in the beginning of each time slot to perform sensing, it can scan for the presence or absence of any PU before transmitting, during the rest of the time slot without colliding with the PU.
If the SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be costly, in terms of energy consumption, especially for wide-band sensing, as detailed in the surveys \cite{yucek2009survey,subhedar2011spectrum}, thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time.
This assumption, along with the hypothesis that PU cannot be disturbed (without which no regulation will never be accepted for OSA), enforces the SU to transmit in the channel that it sensed, if and only if it was sensed free from any PU.
% non negligible cost of radio reconfiguration

Focussing on one SU in an OSA network, it has to sequentially decide a channel to sense (in the set of channels, $[K]=\{1,\dots,K\}$), then it performs sensing, and finally it transmits in this channel if it was sensed free.
The goal of the SU is to minimize its energy consumption (we remind that we focus on ``green'' cognitive radio) and to maximize its uplink data rate, or equivalently, to maximize its number of successful transmission.
%
If the different channels are not uniformly used by the PU, and if we assume a stationary hypothesis on the PU traffic, then the goal of the SU boils down to exploring the different channels and exploiting the best ones.
This frames the Opportunistic Spectrum Access problem as an exploration/exploitation problem,
with a finite set of actions (the channels, also called \emph{arms}),
in a sequential action-then-feedback cycle (time steps are $t\in\N^*$),
under partial information feedback (\ie, the player receives feedback about only one channel over $K$).
These three hypotheses are the ones that restrict from the general ML framework to the specific MAB case (see Figure~\ref{fig:1:ReinforcementLearningCycle}).

@@

We start by formulating the problem studied in this thesis, and then we develop the organization of the contributions.
%
% \textbf{Problematic.}
%
If we sum-up the studied problems and summarize them in one question, it could be the following:
\emph{``Can we adapt the decision making tools already successfully applied to Cognitive Radio for Opportunistic Spectrum Access to the specific needs of CR for the (future) Internet of Things networks?''}
%
We answer partially to this problematic by the following steps.

@@

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core concept is to let every dynamic device run a learning algorithm in a fully decentralized way, in order to optimize the complete system.
It is important to note that each devices communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
It means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic to the other devices following the same objective.
It is well known in game theory that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning lead to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

@@

\item
    \emph{What Doubling-Trick Can and Can't Do for Multi-Armed Bandits},
    by \textbf{Lilian Besson} \& Emilie Kaufmann,
    September $2018$.
    % See Appendix~\ref{app:2:DoublingTricks}.
    Preprint at \href{https://HAL.Inria.fr/hal-01736357}{\texttt{HAL.Inria.fr/hal-01736357}}.
    \cite{Besson2018DoublingTricks}

### 2019-07-07 2-Chapters/2-Chapter/chapter2.tex
Improved all mentions to the (wrong) lower-bound of Ch6
De meilleures transitions, j'ai retravaillé chaque abstract

\abstractStartChapter{}%
%
In this chapter, we present the common base of all the more sophisticated mathematical models studied in this thesis: the stochastic multi-armed bandit (MAB) model, by restricting to the single-player and stationary stochastic case.
%
We motivate this model and discuss its application to and beyond cognitive radio.
We review existing MAB algorithms,
some of which are crucially used in the rest of this thesis, like \UCB, \klUCB{} and Thompson sampling.
%
We focus on decision-making models with a finite number of resources, called arms, and on stochastic models, where an arm is associated with a binary or real-valued distribution.
%
We formalize this MAB model, then we quote some of its important applications.
We define the notion of regret of an algorithm, and present important results from the literature. A short review of the most important families of MAB algorithms is then given.
% We conclude this chapter by discussing the question of online algorithm selection.

@@

\subsection{Regret lower bounds}

@@

>>>>

\abstractStartChapter{}%
%
As explained in Chapter~\ref{chapter:1}, we focus on decision-making models with a finite number of resources, called arms, and on stochastic models, where an arm is associated to a binary or real-valued distribution.
In this chapter, we present the common base of all the more sophisticated mathematical models studied in this thesis: the stochastic multi-armed bandit (MAB) model, by restricting to the single-player and stationary stochastic case.
%
We motivate this model and discuss its application to and beyond cognitive radio.
%
We formalize this MAB model, then we quote some of its important applications.
We define the notion of regret of an algorithm, and present important results from the literature.
A short review of the most important families of MAB algorithms is then given,
and we highlight some algorithms that are crucially used in the rest of this thesis, like \UCB, \klUCB{} and Thompson sampling.
%
% We conclude this chapter by discussing the question of online algorithm selection.

@@

\subsection{Regret lower-bounds}

@@

\begin{figure}[h!]  % [htbp]
	% \centering
	\includegraphics[width=1.12\linewidth]{SP__K5_T10000_N1000__3_algos/main____env1-1_776571591054077529.pdf}
	\caption{Mean regret $R_t$ as function of $t$ for $T=10000$ and $N=1000$, for the 3 most efficient algorithms: $\UCB_1$, \klUCB{} and Thompson Sampling.}
	\label{fig:2:meanRegretAsFunctionOfTimeForDifferentAlgorithmsT10000N1000__FIXME}
\end{figure}

### 2019-07-07 2-Chapters/3-Chapter/chapter3.tex
De meilleures transitions, j'ai retravaillé chaque abstract

\abstractStartChapter{}%
%
SMPyBandits is a Python package I have developed since the beginning of my PhD.
It is designed to allow easy and fast numerical simulations on single-player and multi-players Multi-Armed Bandits (MAB) algorithms.
To the best of the authors' knowledge, it is the most exhaustive open-source implementation of state-of-the-art algorithms and different kinds of MAB models.
%
This chapter details the organization of the library, what it implements in terms of arm distributions, models, algorithms and visualizations.
Then we use it to perform a numerical comparisons of the main state-of-the-art single-player MAB algorithms, and a study to compare time and memory costs of the main algorithms.

@@

$(i)$
Its purpose is to easily implement numerical simulations of stochastic or piece-wise stochastic problems of single- or multi-players multi-armed bandits.
SMPyBandits is distributed on GitHub and Pypi freely, under an open-source license, and it is extensively documented (at \href{https://SMPyBandits.GitHub.io}{\texttt{SMPyBandits.GitHub.io}}).
Our library allows any researcher to easily run numerical simulations of different kinds of multi-armed bandit problems, requiring only a small knowledge of Python thanks to its documentation, its well-designed API, and many examples of simulation scripts and configuration files included.

$(ii)$
We detailed how SMPyBandits is implementing arms, problems, algorithms, and use these components to implement a simulation loop, with various visualizations being performed after the simulation.
As far as now, SMPyBandits is restricted to the finite-arm case, but it supports a wide range of arms distributions.
Different kinds of models are implemented, from stationary single-player to piece-wise stationary multi-players with different collision models.
One of the main qualities of the library is that it is quite exhaustive, as all the main families of algorithms covering these different models have been implemented, even very recent algorithms from the literature, as we followed the active research from December $2016$ to June $2019$.
More than $65$ algorithms or variants of algorithms are implemented for the single-player case, $5$ for the expert aggregation problem, about $15$ for the multi-players case, and about $20$ for the piece-wise stationary problem.
All the codebase is fully documented, and the library is using continuous integration to run automated tests on the code after every modification.
%
When comparing algorithms on a problem, the main performance measure is the regret, but the library also computes, stores and visualizes other measures, such as best-arm selection rate, mean cumulated reward, real time or memory costs.

$(iii)$
Finally, we presented how to use SMPyBandits, if one wants to run some pre-designed simulations or design new simulations.
Running a simulation is very easy, and different examples showed that the main parameters such as the time horizon $T$ or the number of repetitions can be configured directly from the command line when running the Python script, or by modifying the code.
Designing a new simulation requires to have a basic knowledge of Python, but not to dive into the implementation of the library.
We give below in Appendix~\ref{sec:3:appendix} a small but complete example of the two files that has to be written for a new simulation: a \texttt{main} file which essentially defines the desired visualizations after the simulation, and a \texttt{configuration} file which defines the problems to simulate and their parameters, as well as the algorithms that will be compared.

>>>>

\abstractStartChapter{}%
%
SMPyBandits is a Python package I have developed since the beginning of my PhD.
It is designed to allow easy and fast numerical simulations on single-player and multi-players Multi-Armed Bandits (MAB) algorithms.
To the best of the authors' knowledge, it is the most exhaustive open-source implementation of state-of-the-art algorithms and different kinds of MAB models.
%
This chapter details the organization of the library, what it implements in terms of arm distributions, models, algorithms and visualizations.
Then we use it to perform a numerical comparisons of the main state-of-the-art single-player bandit algorithms, and a study to compare time and memory costs of some of the most well-known and most widely used MAB algorithms.

@@

\begin{itemize}
    \item[$(i)$]
Its purpose is to easily implement numerical simulations of stochastic or piece-wise stochastic problems of single- or multi-players multi-armed bandits.
SMPyBandits is distributed on GitHub and Pypi freely, under an open-source license, and it is extensively documented (at \href{https://SMPyBandits.GitHub.io}{\texttt{SMPyBandits.GitHub.io}}).
Our library allows any researcher to easily run numerical simulations of different kinds of multi-armed bandit problems, requiring only a small knowledge of Python thanks to its documentation, its well-designed API, and many examples of simulation scripts and configuration files included.

    \item[$(ii)$]
We detailed how SMPyBandits is implementing arms, problems, algorithms, and use these components to implement a simulation loop, with various visualizations being performed after the simulation.
As far as now, SMPyBandits is restricted to the finite-arm case, but it supports a wide range of arms distributions.
Different kinds of models are implemented, from stationary single-player to piece-wise stationary multi-players with different collision models.
One of the main qualities of the library is that it is quite exhaustive, as all the main families of algorithms covering these different models have been implemented, even very recent algorithms from the literature, as we followed the active research from December $2016$ to June $2019$.
More than $65$ algorithms or variants of algorithms are implemented for the single-player case, $5$ for the expert aggregation problem, about $15$ for the multi-players case, and about $20$ for the piece-wise stationary problem.
All the codebase is fully documented, and the library is using continuous integration to run automated tests on the code after every modification.
%
When comparing algorithms on a problem, the main performance measure is the regret, but the library also computes, stores and visualizes other measures, such as best-arm selection rate, mean cumulated reward, real time or memory costs.

    \item[$(iii)$]
Finally, we presented how to use SMPyBandits, if one wants to run some pre-designed simulations or design new simulations.
Running a simulation is very easy, and different examples showed that the main parameters such as the time horizon $T$ or the number of repetitions can be configured directly from the command line when running the Python script, or by modifying the code.
Designing a new simulation requires to have a basic knowledge of Python, but not to dive into the implementation of the library.
We give below in Appendix~\ref{sec:3:appendix} a small but complete example of the two files that has to be written for a new simulation: a \texttt{main} file which essentially defines the desired visualizations after the simulation, and a \texttt{configuration} file which defines the problems to simulate and their parameters, as well as the algorithms that will be compared.
\end{itemize}

### 2019-07-07 2-Chapters/4-Chapter/chapter4.tex
De meilleures transitions, j'ai retravaillé chaque abstract

\abstractStartChapter{}%
%
In this chapter, we propose two models of Internet of Things (IoT) networks, composed of many independent IoT end-devices,
that can use low-cost RL (Reinforcement Learning) algorithms to learn to improve their spectrum access.
%
Decentralized RL for IoT lets the devices use acknowledgements sent by their Base Station as a reward, instead of sensing feedback like for OSA.
We consider many independent ``dynamic'' devices, communicating with a small probability at every instant.
Simulations show that dynamic devices can improve their spectrum efficiency,
by using MAB algorithms like \UCB.
% in the case of a non-uniform repartition of the static devices in the channels
We also developed a proof-of-concept using USRP platforms, for a real-world validation of this approach.
% Numerical simulations justify the interest of the approach,
%  of the advantage of using MAB algorithms for the first model.
%
% On a simple example, we present in Section~\ref{sec:4:gnuradio} a demonstration that implements and validates the proposed approach, using real wireless radio hardware.
% The first model
% % without retransmission of packets,
% is interesting for its simplicity,
% and because considering no retransmission can improve the battery life of IoT devices, at the cost of a lower Quality of Service (QoS).
%
However, in case of failed transmission of a message, a dynamic device can also retransmit it up-to a fixed number of retransmission.
We compare different heuristics, based on \UCB, and numerical simulations confirm that the non-naive heuristics significantly improve the network efficiency.
% Interestingly, the simplest heuristic is found to be the most efficient one, as its simplicity implies a faster learning and convergence time.
%
% The second model is more interesting if the main target is an improvement of the QoS, rather than a longer battery life.

>>>>

\abstractStartChapter{}%
%
After detailing in Part~\ref{part:Introduction} the MAB model, we now come back to our main question of interest.
In this chapter, we focus on wireless networks following the hypotheses of common to the present and future Internet of Things (IoT) networks exposed in Chapter~\ref{chapter:1}.
Our goal is to show that the IoT devices can automatically learn to increase their battery life as well as their successful transmission rates, without changing anything on the IoT standard.
%
We propose two models of IoT networks, composed of many independent IoT end-devices, that can use low-cost RL (Reinforcement Learning) algorithms to learn to improve their spectrum access.
%
Decentralized RL for IoT lets the devices use acknowledgements sent by their Base Station as a reward, instead of sensing feedback like for OSA.
We consider many independent ``dynamic'' devices, communicating with a small probability at every instant.
Simulations show that dynamic devices can improve their spectrum efficiency,
by using MAB algorithms like \UCB.
% in the case of a non-uniform repartition of the static devices in the channels
We also developed a proof-of-concept using USRP platforms, for a real-world validation of this approach.
% Numerical simulations justify the interest of the approach,
%  of the advantage of using MAB algorithms for the first model.
%
% On a simple example, we present in Section~\ref{sec:4:gnuradio} a demonstration that implements and validates the proposed approach, using real wireless radio hardware.
% The first model
% % without retransmission of packets,
% is interesting for its simplicity,
% and because considering no retransmission can improve the battery life of IoT devices, at the cost of a lower Quality of Service (QoS).
%
However, in case of failed transmission of a message, a dynamic device can also retransmit it up-to a fixed number of retransmission.
We compare different heuristics, based on \UCB, and numerical simulations confirm that the non-naive heuristics significantly also improve the network efficiency.
% Interestingly, the simplest heuristic is found to be the most efficient one, as its simplicity implies a faster learning and convergence time.
%
% The second model is more interesting if the main target is an improvement of the QoS, rather than a longer battery life.

### 2019-07-07 2-Chapters/5-Chapter/ALT_2018__MPBandits.git/mpBandits.tex
De meilleures transitions, j'ai retravaillé chaque abstract

% -----------------------------------------------------------------
% \section{An improved asymptotic regret lower bound (for some algorithms)}
\section[Decomposing the regret to get an intuition about efficient decentralized algorithms]{Decomposing the regret to get an intuition for the design of efficient decentralized algorithms}
\label{sec:5:lowerbound}
% -----------------------------------------------------------------

>>>>

% -----------------------------------------------------------------
% \section{An improved asymptotic regret lower bound (for some algorithms)}
\section[Regret decomposition and intuition about efficient decentralized algorithms]{Decomposing the multi-player regret to get an intuition for the design of efficient decentralized algorithms}
\label{sec:5:lowerbound}
% -----------------------------------------------------------------

### 2019-07-07 2-Chapters/5-Chapter/chapter5.tex
Not sure if I say enough about the lower-bounds...
De meilleures transitions, j'ai retravaillé chaque abstract

\abstractStartChapter{}%
%
In this chapter, we are interested in a more formal approach to the decentralized learning problem presented in Chapter~\ref{chapter:4}.
We restrict to the easier case of at most $M \leq K$ devices in a network with $K$ channels, because it was found very hard to analyze the aforementioned IoT network model.
%
We discuss three feedback levels that give variants of the multi-players MAB model previously studied in the literature.
Using a decomposition of the centralized system regret, we start by explaining the intuition about what an efficient decentralized algorithm should do, and then we propose \RandTopM{} and \MCTopM, two new orthogonalization schemes.
Combining them with the \klUCB{} index policy gives order-optimal algorithms, that achieve state-of-the-art performance,
as illustrated by numerical simulations.
We conclude by reviewing different extensions of the multi-players models.

@@

In all this chapter and the previous research literature, another common hypothesis is that all the $M$ players are pursuing the same goal, and are all behaving nicely by following the same algorithm.
% Efficient utilization of licensed spectrum in the cognitive radio network is challenging due to lack of coordination among the Secondary Users (SUs).
Distributed algorithms proposed in the literature aim to maximize the network throughput by ensuring orthogonal channel allocation for the SUs.
However, these algorithms work under the assumption that all the SUs faithfully follow the algorithms which may not always hold due to the decentralized nature of the network.
In the paper \cite{SawantKumar2018}, the authors study for the first time distributed algorithms that are robust against malicious behavior, also called \emph{jamming attack}.
They consider both the cases of \emph{jammers} launching coordinated and uncoordinated attacks, and consider a set of $J$ jammers.
In the coordinated attack, the jammers select non-overlapping channels to attack in each time slot and can significantly increase the number of collisions for SUs.
They setup the problem in each scenario as a multi-players bandit and develop algorithm,
and their analysis shows that when the SUs faithfully implement proposed algorithms, the regret is constant with high probability.
They validate their claims through exhaustive synthetic experiments and also through a realistic USRP.
Their synthetic experiments consider different cases, for different values of $K$ the number of channels, $M$ the number of players and $J$ the number of jammers.

>>>>

\abstractStartChapter{}%
%
In this chapter, we are interested in a more formal approach to the decentralized learning problem presented in the previous Chapter~\ref{chapter:4}.
We restrict to the easier case of at most $M \leq K$ devices in a network with $K$ channels, that each have data to transmit at every instant, that is they have an activation probability of $p=1$, because it was found very hard to analyze the previously studied IoT network models.
%
We discuss three feedback levels that give variants of the multi-players MAB model previously studied in the literature.
Using a decomposition of the centralized system regret, we start by explaining the intuition about what an efficient decentralized algorithm should do, and then we propose \RandTopM{} and \MCTopM, two new orthogonalization schemes.
Combining them with the \klUCB{} index policy gives order-optimal algorithms, that achieve state-of-the-art performance,
as illustrated by numerical simulations.
We conclude by reviewing different extensions of the multi-players models.

@@

In all this chapter and the previous research literature, another common hypothesis is that all the $M$ players are pursuing the same goal, and are all behaving nicely by following the same algorithm.
% Efficient utilization of licensed spectrum in the cognitive radio network is challenging due to lack of coordination among the Secondary Users (SUs).
Distributed algorithms proposed in the literature aim to maximize the network throughput by ensuring orthogonal channel allocation for the SUs.
However, these algorithms work under the assumption that all the SUs faithfully follow the algorithms which may not always hold due to the decentralized nature of the network.
In the paper \cite{SawantKumar2018}, the authors study for the first time distributed algorithms that are robust against malicious behavior, also called \emph{jamming attack}.
They consider both the cases of \emph{jammers} launching coordinated and uncoordinated attacks, and consider a set of $J$ jammers.
In the coordinated attack, the jammers select non-overlapping channels to attack in each time slot and can significantly increase the number of collisions for SUs.
They setup the problem in each scenario as a multi-players bandit and develop algorithm,
and their analysis shows that when the SUs faithfully implement proposed algorithms, the regret is constant with high probability.
They validate their claims through extensive synthetic experiments and also through a realistic USRP.
Their synthetic experiments consider different cases, for different values of $K$ the number of channels, $M$ the number of players and $J$ the number of jammers.
